---
title: "Fig 6. scATAC-seq uncovers varied transcription initiation in colonic enterocytes."
author: "Liang Huang"
date: "`r Sys.Date()`"
output: html_document
---
(a)-(c) Quality of scATAC data, evaluated by TSS enrichment (a), fragment size (b) and distance from center (c).
```{r setup, include=FALSE}
# devtools::install_github("GreenleafLab/ArchR", ref="master", repos = BiocManager::repositories())

library(Signac)
library(Seurat)
library(ArchR)

setwd("F:/data/GF_mice_scRNAseq/ATAC/Result-X101SC21072094-Z03-J001-B1-37/2.cellranger_atac")

folders <- list.files(pattern = "GF|FMT|SPF") 
inputFiles=gsub("ATAC","ATAC/fragments.tsv.gz",folders)
names(inputFiles) <- gsub("_ATAC","", folders)

addArchRGenome("mm10")
ArrowFiles <- createArrowFiles( inputFiles = inputFiles, sampleNames = names(inputFiles),
                                QCDir ="1.QualityControl",
                                minTSS = 4, minFrags = 1000, addTileMat = TRUE, addGeneScoreMat = TRUE )
ArrowFiles


projHeme1 <- ArchRProject(ArrowFiles = ArrowFiles, outputDirectory ="2.ArchRProject",copyArrows = TRUE)
projHeme1

getAvailableMatrices(projHeme1)
head(projHeme1@cellColData)
head(projHeme1$cellNames)
head(projHeme1$Sample)
quantile(projHeme1$TSSEnrichment)


setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/scATAC")
saveRDS(projHeme1,"./scATAC_ArchRProject.rds")

p1 <- plotGroups(ArchRProj = projHeme1, groupBy ="Sample", colorBy ="cellColData",
                 name ="TSSEnrichment",plotAs ="ridges",baseSize = 10)

ggsave(filename ="TSS_Score.pdf",units = "cm",width = 8, height = 8,plot = p1)


p2 <- plotFragmentSizes(ArchRProj = projHeme1)
p3=p2+theme(legend.title = element_blank(),legend.position = "right")
ggsave(filename ="Fragment_size.pdf",units = "cm",width = 10, height = 8,plot = p3)

p4 <- plotTSSEnrichment(ArchRProj = projHeme1)
p4=p4+theme(legend.title = element_blank(),legend.position = "right")
ggsave(filename ="TSS_profiles.pdf",units = "cm",width = 10, height = 8,plot = p4)


p=p1+p3+p4&theme(text = element_text(size = 8),
  legend.title = element_blank(),
               legend.position = "none")
ggsave(filename ="ATAC_quality.pdf",units = "cm",width = 28, height = 8,plot = p)

```


(d)t-distributed Stochastic Neighbor Embedding (tSNE) of scATAC-seq. 
(e)Ratio of all cell types in different conditions.
```{r setup, include=FALSE}
library(Signac)
library(Seurat)
library(GenomicRanges)
library(future)
library(ggplot2)
library(RColorBrewer)

setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/scATAC")
combined_high_conf=readRDS("./anno_sc_atac.rds")

combined_high_conf=RunTSNE(combined_high_conf,reduction = "pca",dims = 2:30)
n=length(unique(combined_high_conf@meta.data$predicted.id))
combined_high_conf@meta.data$group=factor(combined_high_conf@meta.data$group,levels = c("GF","FMT","SPF"))
value_tsne=as.data.frame(combined_high_conf@reductions[["tsne"]]@cell.embeddings)
minx=min(value_tsne$tSNE_1)
x_limit=max(value_tsne$tSNE_1)-min(value_tsne$tSNE_1)
miny=min(value_tsne$tSNE_2)
y_limit=max(value_tsne$tSNE_2)-min(value_tsne$tSNE_2)

# color=c( "#984EA3","#E41A1C", "#377EB8", "#8DD3C7", "#80B1D3", "#BEBADA", "#FF7F00")
color=brewer.pal(13,"Set2")
DimPlot(combined_high_conf, reduction = "tsne",
        group.by = 'predicted.id',
        cols=colorRampPalette(color)(n),
        split.by = 'group',
        pt.size = 0.1)+
  labs(title="")+
  theme(text = element_text(size=10),
        legend.key.size = unit(0.35,"cm"),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5)
  )+
  geom_segment(aes(x = minx, xend = minx +x_limit/5, y = miny, yend = miny),
               arrow = arrow(length = unit(0.1, "cm")),
               size = 0.5) +
  geom_segment(aes(x = minx, xend = minx, y = miny, yend = miny + x_limit/5),
               arrow = arrow(length = unit(0.1, "cm")),
               size = 0.5)+
  geom_text(aes(x=minx+x_limit/10,y=miny,label="tSNE1"),size=8,size.unit = "pt",vjust=1.5)+
  geom_text(aes(x=minx,y=miny+y_limit/10,label="tSNE2"),size=8,size.unit = "pt",vjust=-1.2,angle=90)+
  coord_cartesian(expand = T,clip = "off")

ggsave("scATAC_tsne.pdf",units = "cm",width = 18,height = 6)


########cell ratio#########
library(ggpie)
library(ggplot2)
library(patchwork)
library(dplyr)

Ratio <- combined_high_conf@meta.data %>%
  group_by(group,predicted.id) %>%
  summarise(n = n()) %>%
  group_by(group) %>%
  mutate(count = n/sum(n)*100)

color=colorRampPalette(color)(n)
names(color)=unique(Ratio$predicted.id)

plot=list()
for (i in c("GF","FMT","SPF")){
  df1=Ratio[Ratio$group==i,]
  leap_cell=unique(Ratio$predicted.id)[!unique(Ratio$predicted.id) %in% df1$predicted.id]
  if(length(leap_cell)>0){
    n=length(leap_cell)
    df1=rbind(df1,data.frame(group=rep(i,n),
                             predicted.id=leap_cell,
                             n=rep(0,n),
                             count=rep(0,n)))
    df1=df1[match(unique(Ratio$predicted.id),df1$predicted.id),]
  }
  plot[[i]] <- ggdonut(df1,
                group_key = "predicted.id",
                donut.label = F,
                label_type = "none",
                count_type = "count",
                border_size = 0,
                fill_color = color,
                r0 = 0.1
                ) +ggtitle(i)
}

plot[[1]]+plot[[2]]+plot[[3]] + plot_layout(guides = "collect") &
  theme(legend.position='right',
        legend.key.size = unit(0.5,"cm"),
        text = element_text(size = 8),
        legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5)) &
  guides(fill = guide_legend(ncol = 2, byrow = TRUE)) 
ggsave("./cell_ratio.pdf",device = cairo_pdf,units = "cm",width = 18,height = 8)
```

(f)Intersection of significantly different peaks. 
(g)Motif significantly enriched in different peaks.
```{r setup, include=FALSE}
library(Signac)
library(Seurat)
library(JASPAR2022)
library(TFBSTools)
library(BSgenome.Mmusculus.UCSC.mm10)
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/scATAC")
# load data
combined_high_conf <- readRDS("./anno_sc_atac.rds")
DefaultAssay(combined_high_conf) <- 'ATAC'

# filter NA and low expression
combined_high_conf <- subset(combined_high_conf, 
                             subset = nCount_ATAC > 1000 & 
                               nCount_ATAC < 50000 &
                               !is.na(nCount_ATAC))

# obtain motif matrix
pfm <- getMatrixSet(
  x = JASPAR2022,
  opts = list(collection = "CORE", tax_group = 'vertebrates', all_versions = FALSE)
)

# add motif
gc()
combined_high_conf=subset(combined_high_conf,predicted.id=="Enterocyte")
combined_high_conf <- AddMotifs(
  object = combined_high_conf,
  genome = BSgenome.Mmusculus.UCSC.mm10,
  pfm = pfm
)

# DE peaks
Idents(combined_high_conf) <- combined_high_conf$group

da_peaks <- FindMarkers(
  object = combined_high_conf,
  ident.1 = 'GF',
  ident.2 = 'SPF',
  only.pos = TRUE,
  test.use = 'wilcox',
  min.pct = 0.05
)
write.csv(file = "./da_peaks_SPF.csv",x = da_peaks)

da_peaks2 <- FindMarkers(
  object = combined_high_conf,
  ident.1 = 'GF',
  ident.2 = 'FMT',
  only.pos = TRUE,
  test.use = 'wilcox',
  min.pct = 0.05
)
write.csv(file = "./da_peaks_FMT.csv",x = da_peaks2)

top.da.peak <- rownames(da_peaks)[da_peaks$p_val_adj < 0.05 & abs(da_peaks$avg_log2FC) > 0.2630344]
top.da.peak2 <- rownames(da_peaks2)[da_peaks2$p_val_adj < 0.05 & abs(da_peaks2$avg_log2FC) > 0.2630344]

########overlap##########
int=list("GF vs SPF"=top.da.peak,"GF vs FMT"=top.da.peak2)
library(venn)
venn(int,
     ilabels = "counts",
     zcolor = "style",
     box=F,
     ggplot = T)
ggsave("./overlap_peaks.pdf",units = "cm",width = 5,height = 5)

com.da.peaks=intersect(top.da.peak,top.da.peak2)

if (length(com.da.peaks) > 500) {
  com.da.peaks <- com.da.peaks[1:500]
}


############# motif enrichment###########
enriched.motifs <- FindMotifs(
  object = combined_high_conf,
  features = com.da.peaks
)

write.csv(enriched.motifs,"./common_DE_peaks_Motif.csv")
#############vocalo plot###########

library(ggplot2)
library(ggrepel)
enriched.motifs=read.csv("./common_DE_peaks_Motif.csv",row.names = 1)

repel_df=enriched.motifs[enriched.motifs$motif.name %in% c("STAT1","FOXP3","Atoh1","Foxo3",
                                                           "GATA4","GATA5","FOXA1","FOXA2"),]
p1=ggplot(enriched.motifs,aes(fold.enrichment,-log2(pvalue),fill=-log2(pvalue)))+
  geom_point(aes(),size=2,shape=21)+
  theme_classic()+
  scale_fill_gradientn(colours = c( "grey", "#c90e1b"))+
  geom_label_repel(data = repel_df,
                   aes(fold.enrichment,-log2(pvalue),label=motif.name),
                  max.overlaps = 100,size=1)+
  labs(title = "Motif of common varied peaks")+
  theme(plot.title=element_text(size=8,hjust = 0.5),
        text=element_text(size = 8),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),legend.position = "none",
        legend.title = element_blank())
p1

ggsave("./NonGF_Motif.pdf",p1,units = "cm",width = 5,height = 5)
```

(h)Different peaks plot of Noct in different conditions. 
```{r setup, include=FALSE}
library(motifmatchr)
library(JASPAR2022)
library(TFBSTools)
library(BSgenome.Mmusculus.UCSC.mm10) 

pfm_info <- getMatrixSet(JASPAR2022, list(collection = "CORE", tax_group = 'vertebrates'))

motif_to_tf <- data.frame(
  motif_id = names(pfm_info),
  tf_name = sapply(pfm_info, function(x) x@name),
  tf_symbol = sapply(pfm_info, function(x) {
    name <- x@name
    symbols <- regmatches(name, gregexpr("[A-Za-z0-9]+", name))[[1]]
    if (length(symbols) > 0) symbols[1] else name
  })
)


gc()
pwm <- getMatrixSet(
  x = JASPAR2022,
  opts = list(species = 10090, all_versions = FALSE)  
)
combined_high_conf <- AddMotifs(combined_high_conf, genome = BSgenome.Mmusculus.UCSC.mm10, pfm = pwm)
gc()

enterocyte=subset(combined_high_conf,predicted.id=="Enterocyte")
Idents(enterocyte)="group"
Idents(enterocyte)=factor(Idents(enterocyte),levels = c( "SPF","FMT","GF"))
enter_anno=Annotation(enterocyte)

regions_highlight <- GRanges(seqnames = "chr3", 
                  ranges = IRanges(start = 50866799, end = 50867719))
CoveragePlot(
  object = enterocyte,
  idents = c("GF", "FMT", "SPF"),
  region = "chr3-50866799-50867719",
  region.highlight = regions_highlight,
  extend.upstream = 1000,
  extend.downstream = 1000
)&theme(text = element_text(size = 8),line = element_line(linewidth = 0.5))
ggsave("Noct_DEpeaks_enterocyte.pdf",units = "cm",width = 10,height = 6)
```

(i)Network which consist of nearest gene of differential peaks position and motif of differential peaks. Foxo3 and Noct are highlighted, and transcription factors of Noct are labeled.
```{r setup, include=FALSE}
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)
library(clusterProfiler)
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/scATAC")

com.da.peaks=read.csv("./common_DE_peaks_Motif.csv",row.names = 1)
peak_list <- strsplit(com.da.peaks, "-")
peaks_gr <- GRanges(
  seqnames = sapply(peak_list, `[`, 1),
  ranges = IRanges(
    start = as.numeric(sapply(peak_list, `[`, 2)),
    end = as.numeric(sapply(peak_list, `[`, 3))
  )
)

peakAnno <- annotatePeak(peaks_gr, 
                         tssRegion = c(-3000, 3000),
                         TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene,
                         annoDb = "org.Mm.eg.db")

anno_df <- as.data.frame(peakAnno)
head(anno_df)
pfm_info <- getMatrixSet(JASPAR2022, list(collection = "CORE", tax_group = 'vertebrates'))

motif_to_tf <- data.frame(
  motif_id = names(pfm_info),
  tf_name = sapply(pfm_info, function(x) x@name),
  tf_symbol = sapply(pfm_info, function(x) {
    name <- x@name
    symbols <- regmatches(name, gregexpr("[A-Za-z0-9]+", name))[[1]]
    if (length(symbols) > 0) symbols[1] else name
  })
)

enriched.motifs.with.tf <- merge(enriched.motifs, motif_to_tf, 
                                 by.x = "motif", by.y = "motif_id", 
                                 all.x = TRUE)

motif.positions <- Motifs(combined_high_conf)

find_peaks_with_motifs <- function(peaks_gr, combined_high_conf, enriched_motifs, n_top = 10) {
  top_motifs=enriched_motifs$motif[order(enriched_motifs$pvalue)]
  result <- list()
  
  motif_data <- GetMotifData(combined_high_conf, assay = "ATAC", slot = "data")
  motif_positions <- GetMotifData(combined_high_conf, assay = "ATAC", slot = "positions")
  
  all_peaks_gr <- StringToGRanges(rownames(combined_high_conf))
  
  for (motif_id in top_motifs) {
    motif_col <- which(colnames(motif_data) == motif_id)
    
    if (length(motif_col) == 0) next
    
    motif_peaks <- which(motif_data[, motif_col] > 0)
    
    if (length(motif_peaks) > 0) {      motif_peaks_gr <- all_peaks_gr[motif_peaks]
      
      overlaps <- findOverlaps(peaks_gr, motif_peaks_gr)
      
      if (length(overlaps) > 0) {
        peak_indices <- queryHits(overlaps)
        
        result[[motif_id]] <- list(
          peaks = peaks_gr[peak_indices],
          motif_name = enriched_motifs$tf_name[enriched_motifs$motif == motif_id],
          motif_symbol = enriched_motifs$tf_symbol[enriched_motifs$motif == motif_id]
        )
      }
    }
  }
  return(result)
}

motif_peak_association <- find_peaks_with_motifs(peaks_gr, combined_high_conf, enriched.motifs.with.tf)

create_final_table <- function(motif_peak_association, anno_df) {
  results <- data.frame()
  
  for (motif_id in names(motif_peak_association)) {
    association <- motif_peak_association[[motif_id]]
    
    for (i in seq_along(association$peaks)) {
      peak <- association$peaks[i]
      peak_id <- paste0(seqnames(peak), "-", start(peak), "-", end(peak))
      
      peak_anno <- anno_df[anno_df$seqnames == as.character(seqnames(peak)) & 
                             anno_df$start == start(peak) & 
                             anno_df$end == end(peak), ]
      
      if (nrow(peak_anno) > 0) {
        result_row <- data.frame(
          peak_id = peak_id,
          motif_id = motif_id,
          tf_name = association$motif_name,
          tf_symbol = association$motif_symbol,
          gene_id = ifelse(!is.na(peak_anno$geneId[1]), peak_anno$geneId[1], NA),
          gene_symbol = ifelse(!is.na(peak_anno$SYMBOL[1]), peak_anno$SYMBOL[1], NA),
          distance_to_tss = ifelse(!is.na(peak_anno$distanceToTSS[1]), peak_anno$distanceToTSS[1], NA),
          annotation = ifelse(!is.na(peak_anno$annotation[1]), peak_anno$annotation[1], NA)
        )
        results <- rbind(results, result_row)
      }
    }
  }
  return(results)
}

final_results <- create_final_table(motif_peak_association, anno_df)
write.csv(final_results,"./motif_target_Genes.csv")






library(igraph)
library(ggraph)
library(ggplot2)
library(dplyr)
library(scales)
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/scATAC")
final_results=read.csv("./motif_target_Genes.csv",row.names = 1)

edges <- final_results %>%
  count(tf_symbol, gene_symbol, name = "weight") %>%
  rename(from = tf_symbol, to = gene_symbol) %>%
  arrange(desc(weight))

nodes <- data.frame(
  name = unique(c(final_results$tf_symbol, final_results$gene_symbol)),
  type = ifelse(unique(c(final_results$tf_symbol, final_results$gene_symbol)) %in% 
                  final_results$tf_symbol, "TF", "Gene"),
  special = ifelse(unique(c(final_results$tf_symbol, final_results$gene_symbol)) %in% 
                     c("Foxo3", "Noct",final_results$tf_symbol[final_results$gene_symbol=="Noct"]), "Highlight", "Normal"),
  stringsAsFactors = FALSE
) %>%
  mutate(
    size = case_when(
      type == "TF" ~ 5,
      TRUE ~ 3
    )
  )

g <- graph_from_data_frame(edges, directed = TRUE, vertices = nodes)

tf_color <- "#1f77b4" 
gene_color <- "#ff7f0e" 
highlight_color <- "#d62728" 

V(g)$color <- case_when(
  V(g)$type == "TF" ~ tf_color,
  TRUE ~ gene_color
)

V(g)$frame.color <- ifelse(V(g)$special == "Highlight", "#d62728", V(g)$color)
V(g)$frame.width <- ifelse(V(g)$special == "Highlight", 1, 1)
E(g)$width <- scales::rescale(edges$weight, to = c(0.5, 3))
E(g)$color <- "gray80"
E(g)$arrow.size <- 0.3

set.seed(123)
plot(g,
     layout = layout_with_fr(g),
     vertex.label = ifelse(V(g)$special == "Highlight", V(g)$name, NA), 
     vertex.label.color = "black",
     vertex.label.cex = 0.8,
     vertex.label.font = 2, 
     main = "TF-Gene Network (Foxo3 and Noct Highlighted)")

legend("bottomright",
       legend = c("Transcription Factor", "Gene", "Foxo3/Noct"),
       pch = 21,
       pt.bg = c(tf_color, gene_color, highlight_color),
       pt.cex = c(1.5, 1.5, 2),
       bty = "n",
       cex = 0.8)

p <- ggraph(g, layout = "kk") +
  geom_edge_link(aes(width = weight, alpha = weight),
                 color = "grey",
                 arrow = arrow(length = unit(2, 'mm'), type = "closed")) +
  geom_node_point(aes(color = type, size = size), alpha = 0.7,
                  data = . %>% filter(special == "Normal")) +
  geom_node_point(aes(fill = type,size = size),
                  color = highlight_color,
                  alpha=0.7,
                  shape = 21, 
                  stroke = 1.5, 
                  data = . %>% filter(special == "Highlight"),show.legend = F) +
  geom_node_label(aes(label = name, filter = special == "Highlight"),
                  repel = TRUE,
                  size = 3,
                  fill = alpha("white", 0.8),
                  label.size = 0.2) +
  scale_edge_width(range = c(0.3, 1.5)) +
  scale_edge_alpha(range = c(0.3, 0.8)) +
  scale_color_manual(values = c("TF" = tf_color, "Gene" = gene_color)) +
  scale_fill_manual(values = c("TF" = tf_color, "Gene" = gene_color)) +
  scale_size_identity() +  
  theme_void() +
  labs(title = "Transcription Factor - Gene Regulatory Network",
       color = "Node Type") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")


ggsave("./network.pdf",p,units = "cm",width = 20,height = 20)
```