---
title: "Fig2.Tissue-specific disturbed genes atlas."
author: "Liang Huang"
date: "`r Sys.Date()`"
output: html_document
---
#a-c Linear Mixed-Effects Model (LMM) screen tissue-specific effector.
(a)Comprehensive evaluation of the linear mixed-effects model, including the distribution of residual values and the ratio of predicted to observed values.
(b)R square distribution of genes.
(c)Screening result according to LMM.
```{r setup, include=FALSE}
rm(list = ls())
library(lme4)
library(lmerTest)
library(tidyverse)
library(broom)
library(car) 
library(future)
library(furrr)

setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/")
data=read.csv("./mRNA_rpkm_symbol.csv",row.names = 1)
data=data %>%
  filter(rowSums(data >0)>=5)
data=as.data.frame(t(data))
group=read.csv("./group.csv")
names(group)[1]="sample_id"
names(group)[2]="treatment"
rownames(group)=group[,1]
group$mouse_id=rep(c(1:9),6)

df=cbind(group,data)
plan(multisession, workers = parallel::detectCores() - 1)

df <- df %>%
  mutate(
    treatment = factor(treatment, levels = c("GF", "FMT", "SPF")),
    tissue = factor(tissue)
  )

gene_names <- names(df)[5:ncol(df)]
fit_model_for_gene <- function(gene_name, data) {
  gene_safe <- paste0("`", gene_name, "`")
  formula <- as.formula(paste(gene_safe, "~ treatment * tissue"))
  model <- tryCatch({
    lm(formula, data = data)
  }, error = function(e) {
    message("Error fitting model for gene: ", gene_name)
    message("Error: ", e$message)
    return(NULL)
  })
  if (is.null(model)) {
    return(tibble(gene = gene_name,
                  term = NA,
                  p.value = NA,
                  statistic = NA,
                  estimate = NA))
  }
  anova_results <- tryCatch({
    car::Anova(model, type = 3)
  }, error = function(e) {
    message("Error in ANOVA for gene: ", gene_name)
    message("Error: ", e$message)
    return(NULL)
  })
  
  if (is.null(anova_results)) {
    return(tibble(gene = gene_name,
                  term = NA,
                  p.value = NA,
                  statistic = NA,
                  estimate = NA))
  }
  interaction_row <- which(rownames(anova_results) == "treatment:tissue")
  
  if (length(interaction_row) == 0) {
    return(tibble(gene = gene_name,
                  term = "treatment:tissue",
                  p.value = NA))
  }
  
  p_value <- anova_results[interaction_row, "Pr(>F)"]
  
  return(tibble(
    gene = gene_name,
    term = "treatment:tissue",
    p.value = p_value
  ))
}
results <- future_map_dfr(
  .x = gene_names,
  .f = ~ fit_model_for_gene(.x, data = df),
  .progress = TRUE
)
plan(sequential)
results <- results %>%
  mutate(p.adj = p.adjust(p.value, method = "fdr"))
significant_genes <- results %>%
  filter(p.adj < 0.05 & !is.na(p.adj)) %>%
  arrange(p.adj)
head(significant_genes)
write.csv(significant_genes, "./machine_learning/significant_genes_interaction_effect.csv", row.names = FALSE)

library(ggplot2)
library(ggrepel)
library(performance)
library(see)
library(lme4)
library(car)
library(emmeans)
library(dplyr)
library(tidyr)
library(patchwork) 
df <- df %>%
  mutate(across(where(is.numeric), ~ log1p(.x))) 
clean_gene_names <- function(names) {
  cleaned <- gsub("[-+/() :]", ".", names)
  cleaned <- ifelse(grepl("^[0-9]", cleaned), paste0("X", cleaned), cleaned)
  return(cleaned)
}
original_names <- names(df)
cleaned_names <- clean_gene_names(original_names)
name_mapping <- data.frame(original = original_names, cleaned = cleaned_names)
names(df) <- cleaned_names
significant_genes$cleaned_gene <- clean_gene_names(significant_genes$gene)
get_model_predictions <- function(gene, data) {
  formula <- as.formula(paste(gene, "~ treatment * tissue"))
  model <- lm(formula, data = data)
  predicted <- predict(model)
  observed <- model$model[[gene]]
  
  return(data.frame(
    gene = gene,
    observed = observed,
    predicted = predicted,
    residual = observed - predicted
  ))
}
n_top <- nrow(significant_genes)
top_genes <- significant_genes$cleaned_gene[1:n_top]
plan(multisession, workers = parallel::detectCores() - 1)
all_predictions <- future_map_dfr(
  .x = top_genes,
  .f = ~ get_model_predictions(.x, df),
  .progress = TRUE
)
plan(sequential)
cat("=== LINEAR MODEL OVERALL EVALUATION REPORT ===\n\n")
overall_r2 <- cor(all_predictions$observed, all_predictions$predicted)^2
cat(sprintf("Overall R-squared: %.4f\n", overall_r2))
mae <- mean(abs(all_predictions$residual))
rmse <- sqrt(mean(all_predictions$residual^2))
cat(sprintf("MAE: %.4f\n", mae))
cat(sprintf("RMSE: %.4f\n", rmse))

{
  bins=50
  p1 <- ggplot(all_predictions, aes(x = observed, y = predicted)) +
    geom_hex(bins = bins) +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
    labs(title = "Predicted vs Observed Values",
         x = "Observed Expression",
         y = "Predicted Expression",
         fill = "Density"
         ) +
    scale_fill_viridis_c(trans = "log") +
    theme_minimal() +
    theme(text = element_text(size = 8),
          plot.title = element_text(size = 8,hjust = 0.5),
          legend.position = "none",
          legend.key.size = unit(0.5,"cm"),
          panel.border = element_rect(fill = NULL))
  
  p2 <- ggplot(all_predictions, aes(x = residual)) +
    geom_histogram(aes(y = ..density..), bins = 100, fill = "steelblue", alpha = 0.7) +
    geom_density(color = "darkred", linewidth = 0.8) +
    scale_x_continuous(limits = c(-0.2,0.2))+
    labs(title = "Residual Distribution",
         x = "Residuals",
         y = "Density") +
    theme_minimal()+
    theme(text = element_text(size = 8),
          plot.title = element_text(size = 8,hjust = 0.5),
          panel.border = element_rect(fill = NULL))
  
  p3 <- ggplot(all_predictions, aes(x = observed, y = residual)) +
    geom_hex(bins = bins) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
    labs(title = "Residuals vs Observed Values",
         x = "Observed Expression",
         y = "Residuals",
         fill = "Density") +
    scale_fill_viridis_c(trans = "log") +
    theme_minimal() +
    theme(text = element_text(size = 8),
          legend.position = "none",
          legend.key.size = unit(0.5,"cm"),
          plot.title = element_text(size = 8,hjust = 0.5),
          panel.border = element_rect(fill = NULL))

    p4 <- ggplot(all_predictions, aes(x = predicted, y = residual)) +
    geom_hex(bins = bins) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
    labs(title = "Residuals vs Predicted Values",
         x = "Predicted Expression",
         y = "Residuals",
         fill = "Density") +
    scale_fill_viridis_c(trans = "log") +
    theme_minimal() +
    theme(text = element_text(size = 8),
          plot.title = element_text(size = 8,hjust = 0.5),
          legend.position = "none",
          legend.key.size = unit(0.5,"cm"),
          panel.border = element_rect(fill = NULL))

  combined_plot <- (p2 + p1) / (p3 + p4) +
    plot_layout(guides = "collect")+
    plot_annotation(title = "Linear Model Overall Evaluation",
                    theme = theme(text = element_text(size = 8),
                                  plot.title = element_text(hjust = 0.5),
                                  plot.subtitle = element_text(hjust = 0.5)),
                    subtitle = sprintf("Overall R² = %.3f, MAE = %.3f, RMSE = %.3f",
                                       overall_r2, mae, rmse))&
    theme(legend.position = "bottom",
          legend.key.width = unit(1.5,"cm"),
          legend.key.height = unit(0.3,"cm"),
          legend.title.position = "top")
  
  pdf("./machine_learning/overall_model_evaluation.pdf", width = 10/2.53, height = 12/2.53)
  print(combined_plot)
  dev.off()
}
gene_r2 <- all_predictions %>%
  group_by(gene) %>%
  summarise(r2 = cor(observed, predicted)^2)

r2_distribution <- ggplot(gene_r2, aes(x = r2)) +
  geom_histogram(bins = 50) +
  geom_vline(aes(xintercept = median(r2)), color = "red", linetype = "dashed") +
  labs(title = "Distribution of R² per Gene",
       x = "",
       y = "Number of Genes") +
  theme_minimal()+
  theme(text = element_text(size = 8),
        panel.border = element_rect(fill = NULL))


pdf("./machine_learning/r2_distribution_per_gene.pdf", width = 10/2.53, height = 4/2.53)
print(r2_distribution)
dev.off()

overall_evaluation <- data.frame(
  Metric = c("R-squared", "MAE", "RMSE"),
  Value = c(overall_r2, mae, rmse)
)

write.csv(overall_evaluation, "./machine_learning/overall_model_performance.csv", row.names = FALSE)
write.csv(gene_r2, "./machine_learning/gene_level_r2.csv", row.names = FALSE)
```

(d)Complex heatmap recover tissue-specific effect genes and pathways using mfuzz clustering.
```{r setup, include=FALSE}
rm(list=ls())
library(ClusterGVis)
library(org.Mm.eg.db)
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/machine_learning")
gene_select=read.csv("./significant_genes_interaction_effect.csv")

data=read.csv("../mRNA_rpkm_symbol.csv",row.names = 1)
df=data[rownames(data)%in%gene_select$gene,]
group=read.csv("../group.csv",row.names = 1)

condition="SPF"
temp_g=group[group$group==condition,]
temp_data=df[,colnames(df)%in%rownames(temp_g)]
tissue_means <- sapply(1:6, function(i) {
  rowMeans(temp_data[, (3 * (i - 1) + 1):(3 * i)])
})
colnames(tissue_means) <- unique(group$tissue)

set.seed(123)
ck=clusterData(tissue_means,
               cluster.method = "mfuzz",
               cluster.num = 6)
visCluster(object=ck,
           ht.col.list = list(col_range=c(-2,-1,0,1,2),
                              col_color=c("#112F61","#4393C4","#F8F8F8","#D6604C","#670920")),
           sample.col = c( "colon" = "#FF6B6B", 
                           "heart" = "#4ECDC4",
                           "jejunum" = "#45B7D1",
                           "kidney" = "#96CEB4",
                           "liver" = "#FFEEAD",
                           "lung" = "#D4A5A5"),
           go.col = rep(jjAnno::useMyCol("calm",n = 6),each = 5),
           cluster.order = c(1:6),
           plot.type="both",
           line.side = "left",
           column_names_rot=45
)
enrich.go=enrichCluster(object = ck,
                        OrgDb = org.Mm.eg.db,
                        type = "BP",
                        pvalueCutoff = 0.05,
                        topn = 5,
                        seed = 5201314)




pdf("./ClusterGVis.pdf",width = 25/2.53,height = 20/2.53)
visCluster(object=ck,
           ht.col.list = list(col_range=c(-2,-1,0,1,2),
                              col_color=c("#112F61","#4393C4","#F8F8F8","#D6604C","#670920")),
           sample.col = c( "colon" = "#FF6B6B", 
                           "heart" = "#4ECDC4",
                           "jejunum" = "#45B7D1",
                           "kidney" = "#96CEB4",
                           "liver" = "#FFEEAD",
                           "lung" = "#D4A5A5"),
           annoTerm.data = enrich.go,
           sample.order = c(1,5,3,2,6,4),
           cluster.order = c(2,5,4,6,3,1),
           go.col = rep(jjAnno::useMyCol("calm",n = 6),each = 5),
           plot.type="both",
           line.side = "left",
           column_names_rot=45
           )
dev.off()

write.csv(ck$long.res,"./cluster_genes_long.csv")
write.csv(ck$wide.res,"./cluster_genes_wide.csv")
```