---
title: "Fig 3. Tissue-specific key effect genes and their behavior in tissue cancer."
author: "Liang Huang"
date: "`r Sys.Date()`"
output: html_document
---


(a)a Importance rank of effector genes screened according LASSO model in each tissue, genes were ranked based on mean decrease accuracy according Randomforest model. Top 3 genes in colon, liver, lung and kidney are markered in red color.
```{r setup, include=FALSE}
library(randomForest)
library(ggplot2)
library(dplyr)
library(tidyr)
library(pheatmap)
library(viridis)
library(caret)
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/machine_learning/LASSO")
lasso_results <- read.csv("lasso_selected_genes_by_tissue.csv")

df <- read.csv("../../mRNA_rpkm_symbol.csv", row.names = 1)
df <- df %>% filter(rowSums(df > 0) >= 5)
df <- as.data.frame(t(df))

group <- read.csv("../../group.csv")
names(group)[1] <- "sample_id"
names(group)[2] <- "treatment"
rownames(group) <- group[,1]
group$mouse_id <- rep(1:9, 6)

df <- cbind(group, df)
df$treatment[df$treatment %in% c("FMT", "SPF")] <- "NGF"


{
set.seed(123)

if (!dir.exists("RandomForest")) dir.create("RandomForest")
find_matching_columns <- function(genes, df_columns) {
  matches <- c()
  for (gene in genes) {
    if (gene %in% df_columns) {
      matches <- c(matches, gene)
      next
    }
    
    gene_dot <- gsub("-", ".", gene)
    if (gene_dot %in% df_columns) {
      matches <- c(matches, gene_dot)
      next
    }
    
    partial_match <- grep(gene, df_columns, value = TRUE, fixed = TRUE)
    if (length(partial_match) > 0) {
      matches <- c(matches, partial_match[1])
      cat("Partial match found for", gene, "->", partial_match[1], "\n")
    } else {
      cat("No match found for gene:", gene, "\n")
    }
  }
  return(unique(matches))
}

all_importance <- list()
res=data.frame(Gene="",Importance ="",Tissue="")
res=res[-1,]

for (tissue in c("colon","liver","jejunum","heart","lung","kidney")) {
  cat("Processing tissue:", tissue, "\n")
  
  tissue_genes <- lasso_results$Gene[lasso_results$Tissue == tissue]
  
  cat("Checking gene names in data frame...\n")
  
  valid_genes <- find_matching_columns(tissue_genes, colnames(df))
  
  if (length(valid_genes) == 0) {
    cat("No valid genes found for", tissue, "- skipping\n")
    next
  }
  
  cat("Found", length(valid_genes), "valid genes for", tissue, "\n")
  cat("Genes:", paste(valid_genes, collapse = ", "), "\n")
  
  tissue_data <- df %>% 
    filter(treatment %in% c("GF", "NGF"))
  
  selected_cols <- c("treatment", valid_genes)
  tissue_data <- tissue_data[, colnames(tissue_data) %in% selected_cols]
  
  if (ncol(tissue_data) < 2) {
    cat("Skipping", tissue, "- insufficient features after filtering\n")
    next
  }
  
  tissue_data$treatment <- as.factor(tissue_data$treatment)
    try({
      x <- as.matrix(tissue_data[, -which(colnames(tissue_data) == "treatment")])
      y <- tissue_data$treatment
      
      rf_model <- randomForest(
        x = x,
        y = y,
        ntree = 500,
        importance = TRUE
      )
      
      imp <- importance(rf_model, type = 1)
      
      if (!is.null(rownames(imp))) {
        imp_df <- data.frame(
          Gene = rownames(imp),
          Importance = imp[, "MeanDecreaseAccuracy"],
          Tissue = tissue,
          stringsAsFactors = FALSE
        )
        
        all_importance[[tissue]] <- imp_df
        imp_df=imp_df[order(imp_df$Importance,decreasing = T),]
        res=rbind(res,imp_df)
        
        tissue_dir <- paste0("RandomForest/", tissue)
        if (!dir.exists(tissue_dir)) dir.create(tissue_dir)
        
        imp_df <- imp_df %>% arrange(desc(Importance))
        if (nrow(imp_df) > 0) {
          imp_df$Gene <- factor(imp_df$Gene, levels = imp_df$Gene)
          
          p <- ggplot(imp_df, aes(x = Importance, y = reorder(Gene, Importance))) +
            geom_bar(stat = "identity", fill = "#46aae0", width = 0.7) +
            labs(title = paste(tissue, "Gene Importance"),
                 x = "Importance (Mean Decrease Accuracy)",
                 y = "") +
            theme_minimal() +
            theme(
              panel.grid.major.y = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line.y = element_line(color = "black"),
              plot.title = element_text(hjust = 0.5, size = 10),
              axis.text = element_text(size = 8)
            )
          
          ggsave(paste0(tissue_dir, "/", tissue, "_importance.pdf"), p, 
                 width = 12/2.53, height = max(4, nrow(imp_df)*0.3)/2.53)
        }
        
        write.csv(imp_df, paste0(tissue_dir, "/", tissue, "_importance.csv"), row.names = FALSE)
        cat("Alternative approach successful for", tissue, "\n")
      }
    }, silent = TRUE)
}
}

write.csv(res,"./RandomForest/importance_result.csv")

res=read.csv("./RandomForest/importance_result.csv",row.names = 1)
res=res[nrow(res):1,]
res$Gene=factor(res$Gene,levels = res$Gene)
res$Tissue=factor(res$Tissue,levels = unique(res$Tissue))
p <- ggplot(res, aes(x = Gene, y = Importance,fill=Tissue)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = c( "colon" = "#FF6B6B", 
                                "heart" = "#4ECDC4",
                                "jejunum" = "#45B7D1",
                                "kidney" = "#96CEB4",
                                "liver" = "#FFEEAD",
                                "lung" = "#D4A5A5"))+
  labs(
       y = "Mean Decrease Accuracy",
       x = "") +
  guides(fill = guide_legend(nrow = 1))+
  theme(
    text = element_text(size = 8),
    panel.background = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5),
    plot.title = element_text(hjust = 0.5, size = 10),
    legend.position = "top",legend.text = element_text(size = 8),
    legend.title = element_blank(),
    legend.key.size = unit(0.3,"cm"),
    axis.text = element_text(size = 8)
  )
p
ggsave("96_genes_importance.pdf", p,
        height= 6/2.53, width = 24/2.53)
```

(b)Expression level of top 3 genes displayed in a way of radar chart.
```{r setup, include=FALSE}
rm(list=ls())
library(tidyverse)
library(ggradar)
library(scales)
library(patchwork)
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/machine_learning/LASSO")
df <- read.csv("../../mRNA_rpkm_symbol.csv", row.names = 1)
df <- df %>% filter(rowSums(df > 0) >= 5)
int=c("NOCT", "MVK", "FOXO3","RORC", "AKR1C19", "GPCPD1","NR1D1", "TRAM2", "CRY1","FKBP5", "NAMPT", "TTLL12")
int=tolower(int)
substring(int,1,1)=toupper(substring(int,1,1))
df=df[rownames(df)%in%int,]
group <- read.csv("../../group.csv",row.names = 1)
group_df <- group %>%
  rownames_to_column("sample") %>%
  as_tibble()

tidy_data <- df %>%
  rownames_to_column("gene") %>%
  pivot_longer(cols = -gene, 
               names_to = "sample", 
               values_to = "expression") %>%
  left_join(group_df, by = "sample") %>%
  group_by(gene, tissue, group) %>%
  summarise(mean_expression = mean(expression, na.rm = TRUE), 
            .groups = "drop") %>%
  pivot_wider(names_from = tissue, values_from = mean_expression) %>%
  mutate(group = factor(group, levels = c("FMT", "GF", "SPF")))

create_radar_plot <- function(gene_data) {
  plot_data <- gene_data %>% select(-gene)
  
  scaled_data <- plot_data %>%
    mutate(across(-group, ~ rescale(.x, to = c(0, 1))))
  
  ggradar(
    scaled_data,
    values.radar = c("0", "0.5", "1"),
    grid.min = 0, grid.mid = 0.5, grid.max = 1,
    grid.label.size = 0,
    axis.label.size = 3,
    group.line.width = 0.5,
    draw.points = F,
    axis.label.offset = 1.2,
    fill = T,
    fill.alpha = 0.3,
    group.colours = c("GF"="#4e62ab","FMT"="#FF9850", "SPF"="#d6404e"), 
    background.circle.colour = "white",
    gridline.mid.colour = "grey70",
    legend.position = "bottom",
    plot.title = unique(gene_data$gene),
    ) +
    coord_cartesian(expand = T,clip = "off") +
    theme(
      plot.title = element_text(hjust = 0.5, size = 10),
      plot.margin = margin(r=30),
      legend.text = element_text(size = 8),
      axis.text.y = element_blank()
    )
}


genes <- int
tidy_data$group=factor(tidy_data$group,levels = c("GF","FMT","SPF"))
plots <- map(genes, ~ {
  gene_data <- filter(tidy_data, gene == .x)
  create_radar_plot(gene_data)
})

pdf("./radar.pdf",width = 15/2.53,height = 15/2.53)
wrap_plots(plots, ncol = 3,guides = "collect") &
  theme(plot.title = element_text(hjust = 0.5, size = 10),
                                    legend.position = "bottom",
                                    plot.margin = margin(t=5,l=20,r=20,b=5),
                                    legend.key.width = unit(0.5,"cm"))
dev.off()


```
(c) Survival analysis of top 3 genes. Red represents high expression, while green represents low expression.
```{r setup, include=FALSE}
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/machine_learning/LASSO")
library(TCGAbiolinks)
library(survival)
library(survminer)
library(SummarizedExperiment)

GM <- list(
  'TCGA-COAD' = c("NOCT", "MVK", "FOXO3"),
  'TCGA-LIHC' = c("RORC", "AKR1C1", "GPCPD1"), 
  'TCGA-LUAD' = c("NR1D1", "TRAM2", "CRY1"),
  'TCGA-KIRC' = c("FKBP5", "NAMPT", "TTLL12")
)

{
plot=list()
for (pro in names(GM)){
  query <- GDCquery(
    project = pro,
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts"
  )
  
  GDCdownload(query, method = "api")
  
  coad_data <- GDCprepare(
    query = query,
    summarizedExperiment = TRUE
  )
  
  gene_list <- GM[[pro]]
  convert_gene_names <- function(gene_symbols) {
    gene_map <- select(org.Hs.eg.db, 
                       keys = gene_symbols, 
                       columns = c("ENSEMBL"), 
                       keytype = "SYMBOL")
    return(gene_map)
  }
  gene_mapping <- convert_gene_names(gene_list)
  plot_list=list()
  for (temp_gene in gene_list){
    noct_expr <- assay(coad_data[grep(
      gene_mapping$ENSEMBL[gene_mapping$SYMBOL==temp_gene],rownames(coad_data)
    ), ])
    rownames(noct_expr) <- temp_gene
    
    clinical_data <- colData(coad_data)
    
    surv_data <- data.frame(
      patient_barcode = clinical_data$patient,
      NOCT_expr = as.numeric(noct_expr[1, ]),
      vital_status = clinical_data$vital_status,
      days_to_death = clinical_data$days_to_death,
      days_to_last_follow_up = clinical_data$days_to_last_follow_up
    )
    
    surv_data$time <- ifelse(
      surv_data$vital_status == "Dead",
      surv_data$days_to_death,
      surv_data$days_to_last_follow_up
    )
    surv_data$status <- ifelse(surv_data$vital_status == "Dead", 1, 0)
    surv_data <- na.omit(surv_data[, c("NOCT_expr", "time", "status")])
    
    cutoff <- median(surv_data$NOCT_expr)
    surv_data$group <- ifelse(surv_data$NOCT_expr > cutoff, "High", "Low")
    
    surv_obj <- Surv(time = surv_data$time, event = surv_data$status)
    km_fit <- survfit(surv_obj ~ group, data = surv_data)
    logrank_test <- survdiff(surv_obj ~ group, data = surv_data)
    cox_model <- coxph(surv_obj ~ NOCT_expr, data = surv_data)
    plot_list[[temp_gene]]=ggsurvplot(  km_fit,
                                        data = surv_data,
                                        pval = TRUE,
                                        linetype = "solid",
                                        conf.int = F,
                                        risk.table = F,
                                        legend.labs = c("High", "low"),
                                        legend.title = temp_gene,
                                        palette = c("#ff837d","#4ECDC4"),
                                        ggtheme = theme(panel.border = element_rect(fill = NULL),
                                                        axis.title = element_blank(),
                                                        plot.title = element_text(hjust = 0.5),
                                                        text = element_text(size = 8),
                                                        panel.grid = element_blank(),
                                                        panel.background = element_blank()),
                                        xlab = "Time (days)",
                                        ylab = "Overall Survival Probability"
    )
    
  }
  
  
  plot[[pro]] <- lapply(plot_list, function(x) x$plot)
  
}
}
mytheme=theme(plot.title = element_text(hjust = 0.5,
                                        margin = margin(b = 0, unit = "pt")),
              legend.position = "top",
              legend.margin = margin(t=0,b=0),
              plot.margin = margin(t=0,b=20,l=10,r=10))
p1= wrap_plots(plot[[1]], ncol = 3)&
  labs(title=names(plot[1]))&mytheme
p2= wrap_plots(plot[[2]], ncol = 3)&theme(legend.position = "top")&
  labs(title=names(plot[2]))&mytheme
p3= wrap_plots(plot[[3]], ncol = 3)&theme(legend.position = "top")&
  labs(title=names(plot[3]))&mytheme
p4= wrap_plots(plot[[4]], ncol = 3)&theme(legend.position = "top")&
  labs(title=names(plot[4]))&mytheme

combined_plot=wrap_plots(list(p1,p2,p3,p4),
                         nrow=4,
                         # guides = "collect",
                         axes = "collect",
                         axis_titles = "collect_x")

pdf("./survival.pdf",width = 15/2.53,height = 24/2.53)
combined_plot
dev.off()
```