---
title: "Fig 1.Gut microbes disturb multi-tissue transcript."
author: "Liang Huang"
date: "`r Sys.Date()`"
output: html_document
---

(b) Gene expression level (RPKM) distribution of each sample.
```{r setup, include=FALSE}
rm (list=ls())
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物")
DATA=read.csv("./mRNA_rpkm_symbol.csv",row.names = 1)

data=data.frame(c(1:nrow(DATA)))
for (tissue in c("colon","liver","jejunum","heart","lung","kidney")){
  temp_data=DATA[,c(grep(paste0(tissue,"GF"),names(DATA)),
                    grep(paste0(tissue,"FMT"),names(DATA)),
                    grep(paste0(tissue,"SPF"),names(DATA))
  )]
  data=cbind(data,temp_data)
  }
data=data[,-1]

library(dplyr)
library(tidyr)
df=data %>%
  filter(rowSums(data >5)>=5)
df$gene=rownames(df)
df=pivot_longer(df,cols = names(df)[-55],names_to = "sample",values_to = "RPKM")
df$tissue=""
df$group=""
for (tissue in c("colon","liver","jejunum","heart","lung","kidney")){
  
  df$tissue[grep(tissue,df$sample)]=tissue
  df$group[grep(tissue,df$sample)]=gsub(tissue,"",df$sample[grep(tissue,df$sample)])
}
df$group=gsub("\\d+\\.*","",df$group)


df$group=factor(df$group,levels = c("GF","FMT","SPF"))
# df=df[order(df$group),]
df$sample <- factor(df$sample, levels = names(data))
df$tissue=factor(df$tissue,levels = c("colon","liver","jejunum","heart","lung","kidney"))

library(ggplot2)
p1=ggplot(df,aes(sample,log10(RPKM+1),fill=group))+
  stat_boxplot(geom="errorbar",width=0.3,position = position_dodge())+
  geom_boxplot(aes(fill=group),outliers = F,width = 0.5,outlier.color = "black",outlier.size = 0.3)+
  facet_wrap(facets = ~tissue,nrow = 1,scales = "free_x")+
  scale_fill_manual(values = c("GF"="#4e62ab","FMT"="#FF9850","SPF"= "#d6404e"))+
  theme(panel.background = element_blank(), 
        text = element_text(size=8),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_blank(),    
        strip.text = element_text(size = 10),
        legend.position = "right",legend.title=element_blank())
p1
ggsave("./rpkm_stasitic.pdf",units = "cm",width = 24,height = 4)

```


(c) PCA of each tissue displayed distance between both three groups.
```{r setup,include=FALSE}
library(DESeq2)
library(ggplot2)
common_theme <- function() {
  list(
    theme(
      panel.background = element_blank(),
      axis.line = element_line(colour = "black", linewidth = 0.5),
      legend.position = "none",
      legend.title = element_blank(),
      plot.title = element_text(size = 10, hjust = 0.5),
      plot.margin = margin(t = 0, r = 10, b = 0, l = 0, unit = "pt") 
    ),
    coord_fixed(ratio = 1) 
  )
  }
setwd("")
DATA=read.table("./all_mRNA_counts.txt",header = T)
data=DATA[,7:60]
rownames(data)=DATA$Geneid
data=data[rowSums(data)!=0,]
data=as.data.frame(data)
data=data[rowSums(data>0)>3,]
group=read.csv("./group.csv",row.names = 1)

plot=list()
for (i in c("colon","liver","jejunum","heart","kidney", "lung")){
  coldata=group[group$tissue==i,]
  temp_data=data[,colnames(data)%in%rownames(coldata)]
  coldata$group=factor(coldata$group)
  coldata$group <- relevel(coldata$group, ref = "GF")
  dds <- DESeqDataSetFromMatrix(countData = temp_data,
                                colData = coldata,
                                design=  ~group)
  dds <- DESeq(dds)
  vsd <- vst(dds, blind=FALSE)
  p=plotPCA(vsd, intgroup=c("group"))+
    scale_color_manual(values = c("#4e62ab","#FF9850", "#d6404e"))+
    labs(title = i)+
    coord_cartesian(expand = F)+
    common_theme()
  plot[[i]]=p
}
library(patchwork)
pdf("./PCA_6_tissue.pdf",width = 30/2.53,height = 4/2.53)
plot[[1]]+plot[[2]]+plot[[3]]+plot[[4]]+plot[[5]]+plot[[6]]+
  plot_layout(ncol = 6, nrow = 1,guides = "collect")&
  theme(aspect.ratio = 1,
        text = element_text(size=8),
        legend.position = "right")
dev.off()
```


(d)Heatmap of coding and non-coding mRNA relative expression level.
```{r setup,include=FALSE}
rm(list = ls())
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物")
data=read.csv("./mRNA_rpkm_symbol.csv",row.names = 1)
group=read.csv("./group.csv",row.names = 1)
group$group=factor(group$group,levels = c("GF","FMT","SPF"))
anno=read.csv("./annotation.csv",row.names = 1)
setwd("./mRNA/second/")
df_mRNA=data[rownames(data)%in%anno$external_gene_name[anno$gene_biotype=="protein_coding"],]
df_miRNA=data[rownames(data)%in%anno$external_gene_name[anno$gene_biotype=="miRNA"],]
df_lncRNA=data[rownames(data)%in%anno$external_gene_name[anno$gene_biotype=="lncRNA"],]

library(RColorBrewer)
display.brewer.all()
tis=brewer.pal(6, "Accent")
names(tis)=unique(group$tissue)
g=c("GF"="#4e62ab","FMT"="#FF9850", "SPF"="#d6404e")
color=list(group=g,
           tissue=tis)
library(pheatmap)
hmap=function(df,title){
  p=pheatmap(df,kmeans_k = 300,
             scale="row",
             show_colnames = F,
             show_rownames = F,
             annotation_col = group,annotation_colors = color,
             color = colorRampPalette(colors = c("#112F61","#4393C4","#F8F8F8","#D6604C","#670920"))(11),
             main = title,fontsize = 8)
  return(p)
}

p1=hmap(df_mRNA,"mRNA")
p2=hmap(df_lncRNA,"lncRNA")
p3=hmap(df_miRNA,"miRNA")

library(gridExtra)
pdf("./heatmap_all.pdf",width = 25/2.53,height = 8/2.53)
grid.arrange(grobs=list(p1$gtable,p2$gtable,p3$gtable),ncol=3)
dev.off()
```

# d & Fig S1 Screen DEGs and performed KEGG pathway enrichment.
(e)Differentially expressed genes (DEGs) counts in each tissue.
Fig S1. KEGG pathway enrichment of genes disturbed by gut microbes.
```{r setup,include=FALSE}
rm(list = ls())
############input data###############
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物")
DATA=read.table("./all_mRNA_counts.txt",header = T)
names(DATA)=gsub("...bam.","",names(DATA))
names(DATA)=gsub("_Aligned.sortedByCoord.out.bam","",names(DATA))

data=DATA[,7:60]
rownames(data)=DATA$Geneid
data=data[rowSums(data)!=0,]
group=read.csv("./group.csv",row.names = 1)

setwd("./mRNA/")
###########mapID############
library(org.Mm.eg.db)
library(clusterProfiler)
library(DESeq2)
library(tidyverse)

genenames=bitr(rownames(data),fromType = "ENSEMBL",OrgDb = org.Mm.eg.db,toType = "SYMBOL")
genenames=genenames[!duplicated(genenames$ENSEMBL),]
genenames=genenames[!duplicated(genenames$SYMBOL),]
data=data[rownames(data)%in%genenames$ENSEMBL,]
rownames(data)=genenames$SYMBOL[rownames(data)%in%genenames$ENSEMBL]
data=data[rowSums(data)!=0,]
write.csv(data,"./mRNA_symbol.csv")
###############DEGs venn in each tissue##################
{
  colData <- group
  tissues <- unique(colData$tissue)
  all_results <- list()
  for (tissue in tissues) {
    tissue_samples <- rownames(colData[colData$tissue == tissue, ])
    tissue_data <- data[, tissue_samples]
    tissue_colData <- colData[tissue_samples, ]
    dds <- DESeqDataSetFromMatrix(
      countData = tissue_data,
      colData = tissue_colData,
      design = ~ group
    )
    keep <- rowSums(counts(dds) >= 10) >= 1
    dds <- dds[keep, ]
    dds <- DESeq(dds)
    comparisons <- list(
      c("SPF","GF"),
      c("SPF", "FMT"),
      c("FMT", "GF")
    )
    for (comp in comparisons) {
      group1 <- comp[1]
      group2 <- comp[2]
      res <- results(dds, contrast = c("group", group1, group2))
      res_df <- as.data.frame(res)
      res_df$gene <- rownames(res_df)
      res_df$comparison <- paste(group2, "vs", group1)
      res_df$tissue <- tissue
      all_results[[paste(tissue, group2, "vs", group1)]] <- res_df
    }
  }
  final_results <- do.call(rbind, all_results)
}
setwd("./second/")
write.csv(final_results, file = "DEGs_conditions.csv", row.names = FALSE)

################DEGs ###########
setwd("E:/博士2023-2025/课题/跨组织转录组-肠道微生物/mRNA/second/")
rm(list=ls())
library(dplyr)
data=read.csv("./DEGs_conditions.csv",header = T)
DE_data <- data %>%
  filter(pvalue < 0.05 & abs(log2FoldChange) > 0.2630344)

anno=read.csv("../../annotation.csv",row.names = 1)
DE_data$genetype=anno$gene_biotype[match(DE_data$gene,anno$external_gene_name)]
library(ggplot2)
library(clusterProfiler)
library(org.Mm.eg.db)
library(patchwork)
{
DEGs_count=list()
df_count=data.frame()
KEGG=data.frame()
enrich=data.frame()
plot=list()

ov=data.frame(genes=0,type=0,tissue=0)
ov=ov[-1,]
for (tissue in c("colon","liver","jejunum","heart","lung","kidney")){
  temp_data=DE_data[DE_data$tissue==tissue,]
  DEG_list=list()
  ####get DEGs of SPF and FMT
  for (comparison in c("GF vs SPF","GF vs FMT")){
    DEG_list[[comparison]]=list(
      mRNA_up=temp_data$gene[which(temp_data$comparison==comparison & temp_data$genetype=="protein_coding" & temp_data$log2FoldChange > 0)],
      miRNA_up=temp_data$gene[which(temp_data$comparison==comparison & temp_data$genetype=="miRNA" & temp_data$log2FoldChange >0)],
      lncRNA_up=temp_data$gene[which(temp_data$comparison==comparison & temp_data$genetype=="lncRNA"& temp_data$log2FoldChange >0)],
      mRNA_dn=temp_data$gene[which(temp_data$comparison==comparison & temp_data$genetype=="protein_coding"& temp_data$log2FoldChange < 0)],
      miRNA_dn=temp_data$gene[which(temp_data$comparison==comparison & temp_data$genetype=="miRNA"& temp_data$log2FoldChange < 0)],
      lncRNA_dn=temp_data$gene[which(temp_data$comparison==comparison & temp_data$genetype=="lncRNA"& temp_data$log2FoldChange < 0)]
    )
  }
  DEG_list[["overlap"]]=list()
  count=c()
  ########get overlap DEGs between SPF and FMT
  for(type in c("mRNA_up","miRNA_up","lncRNA_up","mRNA_dn","miRNA_dn","lncRNA_dn")){
  overlap=intersect(DEG_list[[1]][[type]],DEG_list[[2]][[type]])
  DEG_list[["overlap"]][[type]]=overlap
  if(length(overlap!=0)){
      temp_ov=data.frame(genes=overlap,type=type,tissue=tissue)
      ov=rbind(ov,temp_ov)
    }  
  count=c(count,
              count=as.numeric(c(length(DEG_list[[1]][[type]]),
                length(DEG_list[[2]][[type]]),
                length(DEG_list[[3]][[type]])
                )
              )
  )
  }
  df=data.frame(
    condition=c(rep(c("SPF","FMT","overlap"),6)),
    genetype=c(rep(c(rep("mRNA",3),rep("miRNA",3),rep("lncRNA",3)),2)),
    regulation=c(rep(c(rep("up",3),rep("down",3)),3)),
    count=count
    )
  df$tissue=tissue
  df_count=rbind(df_count,df)
  ########kegg############
  temp_out=data.frame("category"=0,"subcategory"=0, "ID"=0,"Description"=0, "GeneRatio"=0,
                      "BgRatio"=0,"pvalue"=0,"p.adjust"=0,"qvalue"=0,
                      "geneID"=0,"Count"=0,"tissue"=0,"reg"=0 )
  temp_out=temp_out[-1,]
  for (i in c(1,4)){
    genes=c(DEG_list[[3]][[i]],DEG_list[[3]][[i+1]],DEG_list[[3]][[i+2]])
    df=bitr(genes,"SYMBOL","ENTREZID",OrgDb = org.Mm.eg.db)
    kegg=enrichKEGG(df$ENTREZID,organism = "mmu",
                    pvalueCutoff = 0.05,
                    pAdjustMethod = "BH",
                    qvalueCutoff = 1)
    OUT=kegg@result
    OUT=na.omit(OUT)
    out=OUT
    if(nrow(out)!=0){
      out$Description=gsub("\\ - .*","",out$Description)
      out$tissue=tissue
      out=out[!duplicated(out$Description),]
      enrich=rbind(enrich,out)
      temp=head(out,10)
      temp=temp[order(temp$p.adjust,decreasing = T),]
      temp$reg=ifelse(i==1,"up","down")
      temp_out=rbind(temp_out,temp)
    }
  }
  temp_out=temp_out[!duplicated(temp_out$Description),]
  temp_out$Description=factor(temp_out$Description,levels = unique(temp_out$Description))
  p1=ggplot(temp_out, aes(x = -log10(p.adjust), y = Description))+
    geom_col(aes(fill=reg),alpha=0.3,color="black",width = 0.85)+
    scale_fill_manual(values = c("up"="#d6404e","down"="#3789D8"))+
    coord_cartesian(expand = T,clip = "off") +
    theme(panel.background = element_blank(),
          text = element_text(size = 8),
          plot.title=element_text(hjust = 0.5),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.line.x = element_line(),legend.position = "none",
          plot.margin = margin(t=10,b=10,r = 220,unit = "pt"))+
    geom_text(aes(label = Description), size = 8,size.unit = "pt",hjust=-0.02,vjust=0.5)+
    labs(x="-log10 (padj)",y="", title = tissue)
  p1
  KEGG=rbind(KEGG,temp_out)
  plot[[tissue]]=p1
}
}

write.csv(ov,"./overlap_genes.csv")
write.csv(KEGG,"./overlap_KEGG_result.csv")
library(patchwork)
pdf("./KEGG_overlap.pdf",width = 36/2.53,height = 18/2.53)
plot[[1]]+plot[[2]]+plot[[3]]+plot[[4]]+plot[[5]]+plot[[6]]+
  plot_layout(ncol=3,nrow=2,guides = "collect")&
  theme(legend.position = "bottom",
        legend.title=element_blank())
dev.off()


#########DEGs count plot##########
df_count_bidirectional <- df_count %>%
  mutate(count = ifelse(regulation == "down", -count, count))
df_count_bidirectional$tissue=factor(df_count_bidirectional$tissue,levels = c("colon","liver","jejunum","heart","lung","kidney"))
df_count_bidirectional$condition=factor(df_count_bidirectional$condition,level=c("SPF","FMT","overlap"))
df_count_bidirectional$genetype=factor(df_count_bidirectional$genetype,level=c("miRNA","lncRNA","mRNA"))

pdf("./DEGs_overlap(none).pdf",width = 21/2.53,height = 6/2.53)
ggplot(df_count_bidirectional, 
       aes(x = condition, y = count)) +
  geom_col(aes(fill = genetype),
           position = position_dodge(width = 0.7),
           width = 0.7) +
  geom_text(aes(label = ifelse(count==0,"",abs(count)),
                y = ifelse(count > 0, 
                           count + max(count)*0.02,
                           count - max(abs(count))*0.02),
                color=genetype,
                group = genetype),
            position = position_dodge(width = 0.7),
            vjust = ifelse(df_count_bidirectional$count > 0, -0.5, 1.5),
            size = 6,
            size.unit = "pt") +
  geom_text(data = subset(df_count_bidirectional, tissue == "colon"),
            aes(x = -Inf, y = Inf, label = "up"),
            hjust = -0.1, vjust = 1.5, color = "black", size = 2) +
  geom_text(data = subset(df_count_bidirectional, tissue == "colon"),
            aes(x = -Inf, y = -Inf, label = "down"),
            hjust = -0.1, vjust = -0.5, color = "black", size = 2) +
  scale_fill_manual(values = c("#9841A3", "#ff745e", "#3789D8")) +
  scale_color_manual(values = c("miRNA"="#9841A3","lncRNA" = "#ff745e","mRNA" = "#3789D8")) +
  labs(x = "", y = "DEGs count (compare to GF)") +
  theme(
    text = element_text(size = 8),
    panel.background = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA),
    plot.title = element_text(hjust = 0.5, vjust = 3),
    legend.position = "top", 
    legend.title = element_blank(),
    plot.margin = margin(5, 5, 0, 5)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0.1, 0.1)),
    limits = c(-max(abs(df_count_bidirectional$count)) * 1.1, 
               max(abs(df_count_bidirectional$count)) * 1.1),
    labels = abs
  ) +
  geom_hline(yintercept = 0, color = "black") +
  facet_wrap(vars(tissue), nrow = 1)
dev.off()
```